On the Synchronization Power of the Openflow Data Plane

### 2015-05-21 inband-sync.tex
one more

\newcommand{\error}{\texttt{error}}
\newcommand{\xid}{\texttt{xid}}
\newcommand{\ecode}{\texttt{error-code}}
\newcommand{\bundle}{\textbf{atomic}}
\newcommand{\ack}{\textit{ack}}

@@

More precisely, the match field of a rule is 
a  a ternary pattern over packet header fields
(a value and a mask).
\petr{example?}
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
\petr{unclear what ``sometimes'' means here}
%(For instance, the meta-data field is maskable.)

@@

\paragraph{Interaction between control and data planes: FlowMod commands}
The flow tables installed on the switches across the data plane
determine the network \emph{policy}.
Controllers can change the policy via sending  
control messages containing \emph{FlowMod commands}.
In a nutshell, a \emph{FlowMod} command either specifies a new flow entry or
a modifications to an existing flow entry. \petr{The description below
  does not explain how a flow entry can be modified.}  
A flow entry consists of: (1)~a match 
field
%defined by a ternary pattern over packet header fields
(a value and a mask), (2)~a set of actions, and (3)~a priority.
%,a match and an action.
The standard processing of a FlowMod command received by a switch is
as follows. 
If the switch already maintains a flow entry with \emph{exactly} the
same match, then the new flow entry will simply \emph{replace} it.
\petr{What if the two rules have different priorities?}
Otherwise, the flow entry will be installed in addition to existing ones. 
%but will be installed
%\emph{in addition} to existing entries if they only
%partially overlap (or define flowspace subsets or supersets).

@@

%[[PK not sure this is the right place 
%implement synchronization objects which form
%the basis of our transactional interface.
%]]
We will use the following notation to create a \emph{FlowMod} message:
$\FlowMod(\match, \action, \flags)$, $\match$ denotes the match,
$\action$ the action, and $\flags$ the flags to activate the $\checko$
feature.
\petr{Sometimes we also write something like op=add in flowmod, should
  we specify it here?}  
The \emph{FlowMod} message (and also the flow entries) include more parameters,
which however are irrelevant for our implementation: we suggest to
use the default values.

In order to read the current configuration of a switch (the existing flow entries), the controller should send
%By installing 
the OpenFlow \emph{add flow monitor} command with the \textsf{OFPFMF\_INITIAL} optional flag.
%In reply the switch will send all the flow entries defined in the switch.
In this paper, we indicate sending this message and gathering the responds (the reply messages containing the flow entries) by \emph{read configuration} statement which also returns the configuration.
%, a controller can read the current configuration. 
% %(and subscribe for updates). 
%
%can i match multiple flows in multiple tables? yes, but only for FlowMod, not GroupMod: but that is good enough!%
\liron{I changed a bit the description of flow monitors following Petr's comment, I hope it's more clear now.}
\petr{Not very clear what reading the configuration means: is it done
  in an atomic way?}

@@

Another important feature in Openflow is the \emph{bundle}. 
A controller can send multiple $\FlowMod$ commands  equipped with 
the same \emph{bundle identifier}.
These commands are buffered temporarily by
the switch until a \emph{bundle commit} message is received from the
controller. 
The bundled commands are then performed with \emph{all-or-nothing semantics}:
either all of the commands are performed, or none of them is. 
In particular, if a configuration request contained in one of the bundled commands cannot
be applied (e.g.,  all the command is rejected because of the set
$\checko$ flag and a conflicting flow entry), all  other commands in
the bundle are rejected and an error
message is sent to the controller that issued them.
%If multiple \emph{FlowMod} messages with the $\checko$ flag are sent as a single bundle,
The error message contains $\xid$, the identifier  first failed command in the
bundle, and $ecode$, the \emph{error code} corresponding to the failure.   
%[[PK too detailed and not very clear
%Concretely, when the commit fails, the switch must generate an error message corresponding to the
%message that could not be applied. The error message must use the identifier (\emph{xid}) of the offending message,
%the error data field corresponding to that message and the error code corresponding to the failure. If
%multiple messages are generating errors, the switch may return only the first error found or generate
%multiple error messages for the same bundle.
%]]

%When properly configured,
Bundle execution is considered controller atomic in the sense that all controllers can observe switch states (configurations) that either \emph{before} or \emph{after} the
bundle (i.e., all its commands) are executed. 
By specifying the flag OFPBF\_ATOMIC,  
a bundle can also be  \emph{packet-atomic}, in a similar sense, which can prevent packets to being processed by intermediate (possibly invalid) configuration during a bundle.
In addition, by specifying the flag OFPBF\_ORDERED,  bundle commands are executed according to the order they were added to the bundle, thereby allowing to perform dependent actions inside a bundle.
\liron{i added controller atomicity and order preservation, see that it make sense...}

@@

We distinguish two parts of the configuration space of a switch:
the ``normal'' configuration which determines the network policy and
is used to process the data-plane traffic (e.g., forwarding rules), 
and the ``meta'' configuration that contains information used by the
controllers to synchronize their actions.

\petr{How to make sure that traffic is not affected by the ``meta'' part?}

\subsection{Transactional interface}

Our framework provides the controllers with a transactional interface to the
switch configuration.     
A transaction consists of a sequence of \emph{operations}, which
should either be executed in its entirety or completely rejected. 
An operation can be a regular FlowMod command or a synchronization
primitive, which, as we
show below translate into specific FlowMod commands operating on the
switch meta-configuration.

We will use the notation
%$\textbf{bundle}\{\;\ldots\;\}\rightarrow res$ to refer to represent a
$\bundle\{op_1,\;\ldots\;op_k\}$ to refer to a
sequence of commands $op_1,\ldots,op_k$ sent in a bundle.
%\petr{What about replacing $\top-\bot$ with ack-nack? Can we also get a reason why a bundle did not commit?}
The bundle construct is implemented as a sequence of the following
control messages:
\begin{enumerate}
\item A \emph{create bundle} message: a \texttt{ofpt\_bundle\_control} message of type
\texttt{ofpbct\_open\_request} equipped with \textit{bid}, the bundle identifier.
\item A sequence of commands, each wrapped in an
\emph{add to bundle message} (\texttt{ofpt\_bundle\_add\_message})
equipped with \emph{bid}.
\item Finally, a \emph{commit bundle} control message (\texttt{ofpt\_bundle\_control}) of
type \texttt{ofpbct\_commit\_request} with \textit{bid}.
\end{enumerate}
The construct returns $\ack$ if the bundle \emph{commits} and all commands are performed successfully, and
$\error(\xid,\ecode)$ received from the switch otherwise.
%Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).
\petr{Should also provide the reason why the bundle has failed?}

>>>>

\newcommand{\error}{\texttt{error}}
\newcommand{\xid}{\texttt{xid}}
\newcommand{\ecode}{\texttt{error-code}}
\newcommand{\exec}{\textbf{execute}}
\newcommand{\execatomic}{\textbf{execute-atomic}}
\newcommand{\ack}{\textit{ack}}

@@

More precisely, the match field of a rule is 
a ternary pattern over packet header fields
(a value and a mask).
\petr{example?}
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
\petr{unclear what ``sometimes'' means here}
%(For instance, the meta-data field is maskable.)

@@

\paragraph{Interaction between control and data planes: FlowMod commands}
The flow tables installed on the switches across the data plane
determine the network \emph{policy}.
Controllers can change the policy via sending  
control messages containing \emph{FlowMod commands}.
In a nutshell, a \emph{FlowMod} command either specifies a new flow entry or
a modification to an existing flow entry. \petr{The description below
  does not explain how a flow entry can be modified.}  
A flow entry consists of: (1)~a match 
field
%defined by a ternary pattern over packet header fields
(a value and a mask), (2)~a set of actions, and (3)~a priority.
%,a match and an action.

A FlowMod command can either add a flow entry or delete 
The standard processing of a FlowMod $\add$ command received by a switch is
as follows. 
If the switch already maintains a flow entry with \emph{exactly} the
same match and priority level, then the new flow entry will simply \emph{replace} it.
\petr{What if the two rules have different priorities?}
Otherwise, the flow entry will be installed in addition to existing
ones.
A $\dele$ command simply remove an existing flow entry with the same
match and priority  or does nothing if such an entry is not present.
\petr{Is this correct?}

%but will be installed
%\emph{in addition} to existing entries if they only
%partially overlap (or define flowspace subsets or supersets).

@@

%[[PK not sure this is the right place 
%implement synchronization objects which form
%the basis of our transactional interface.
%]]
We will use the following notation to create a \emph{FlowMod} message:
$\FlowMod(\match, \add/\dele, \action, \flags)$, where the first field
$\match$ denotes the match,
the second introduces 
$\action$ the action, and $\flags$ the flags to activate the $\checko$
feature.

\petr{Sometimes we also write something like op=add in flowmod, should
  we specify it here?}  
The \emph{FlowMod} message (and also the flow entries) include more parameters,
which however are irrelevant for our implementation: we suggest to
use the default values.

In order to read the current configuration of a switch (the existing flow entries), the controller should send
%By installing 
the OpenFlow \emph{add flow monitor} command with the \textsf{ofpt\_initial} optional flag.
%In reply the switch will send all the flow entries defined in the switch.
In this paper, we indicate sending this message and gathering the responds (the reply messages containing the flow entries) by \emph{read configuration} statement which also returns the configuration.
%, a controller can read the current configuration. 
% %(and subscribe for updates). 
%
%can i match multiple flows in multiple tables? yes, but only for FlowMod, not GroupMod: but that is good enough!%
\liron{I changed a bit the description of flow monitors following Petr's comment, I hope it's more clear now.}
\petr{Not very clear what reading the configuration means: is it done
  in an atomic way?}

@@

Another important feature in Openflow is the \emph{bundle}. 
A controller can send multiple $\FlowMod$ commands  equipped with 
the same \emph{bundle identifier}.
These commands are buffered temporarily by
the switch until a \emph{bundle commit} message is received from the
controller. 
The bundled commands are then performed with \emph{all-or-nothing semantics}:
either all of the commands are performed, or none of them is. 
In particular, if a configuration request contained in one of the bundled commands cannot
be applied (e.g.,  all the command is rejected because of the set
$\checko$ flag and a conflicting flow entry), all  other commands in
the bundle are rejected and an error
message is sent to the controller that issued them.
%If multiple \emph{FlowMod} messages with the $\checko$ flag are sent as a single bundle,
The error message contains $\xid$, the identifier  first failed command in the
bundle, and $\ecode$, the \emph{error code} corresponding to the failure.   
%[[PK too detailed and not very clear
%Concretely, when the commit fails, the switch must generate an error message corresponding to the
%message that could not be applied. The error message must use the identifier (\emph{xid}) of the offending message,
%the error data field corresponding to that message and the error code corresponding to the failure. If
%multiple messages are generating errors, the switch may return only the first error found or generate
%multiple error messages for the same bundle.
%]]

The bundles used in this paper will always have the following structure:
\begin{enumerate}
\item First, a bundle contains a \emph{create bundle} message: a \texttt{ofpt\_bundle\_control} message of type
\texttt{ofpbct\_open\_request}.
\item The follows a sequence of commands (e.g., \emph{FlowMod}), each wrapped in an
\emph{add to bundle message} (\texttt{ofpt\_bundle\_add\_message}) with the \emph{bundle identifier}.
\item Finally, a \emph{commit bundle} control message (\texttt{ofpt\_bundle\_control}) of
type \texttt{ofpbct\_commit\_request}.
\end{enumerate}
Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).

%When properly configured,
Bundle execution is considered controller atomic in the sense that all controllers can observe switch states (configurations) that either \emph{before} or \emph{after} the
bundle (i.e., all its commands) are executed. 
By specifying the flag \texttt{OFPBF\_ATOMIC},  
a bundle can also be  \emph{packet-atomic}, in a similar sense, which can prevent packets to being processed by intermediate (possibly invalid) configuration during a bundle.
In addition, by specifying the flag \texttt{OFPBF\_ORDERED},  bundle commands are executed according to the order they were added to the bundle, thereby allowing to perform dependent actions inside a bundle.
\liron{i added controller atomicity and order preservation, see that it make sense...}

@@

We distinguish two parts of the configuration space of a switch:
the ``normal'' configuration which determines the network policy and
is used to process the data-plane traffic (e.g., forwarding rules), 
and the \emph{meta-configuration} that contains information used by the
controllers to synchronize their actions.

The meta-configuration is implemented as a set of flow entries
designed in a way that it does not affect the processing of data-plane
traffic. One way to do it is to set a low priority to
the meta flow entries and ensure that the regular flow entries
maintain higher priority levels.
Alternatively, we could leverage the standard ability of maintaining multiple
flow tables at an OpenFlow switch, and use a separate flow table for
the meta-configuration,  making sure that the
\emph{pipelining} mechanism~\cite{of-spec} will never involve this
flow table.

%\petr{How to make sure that traffic is not affected by the ``meta'' part?}

\subsection{Synchronization}

Our framework provides the controllers with the ability to execute
sequences of \emph{operations} on the 
switch configuration in an \emph{atomic} (transactional) or \emph{non-atomic} mode.     
In the atomic mode, all operations the sequence (which we also refer
to as a \emph{transaction}) 
should either be executed in its entirety or completely rejected. 
In the non-atomic mode, no atomicity guarantees are provided. 
An operation here can be a regular FlowMod command or a synchronization
primitive, which, as we
show below, translates into a sequence of FlowMod commands operating on the
switch meta-configuration.

We propose the following constructs
%$\textbf{bundle}\{\;\ldots\;\}\rightarrow res$ to refer to represent a
$\exec\{op_1,\;\ldots\;op_k\}$ and $\execatomic\{op_1,\;\ldots\;op_k\}$ 
which refer to executing a sequence of commands $op_1,\ldots,op_k$ in
a regular and atomic way, respectively.
The construct returns $\ack$ if the sequence of operations is performed and all commands are performed successfully, and
$\error(\xid,\ecode)$ received from the switch otherwise, providing
the identifier of the first command in the sequence that failed
together with the corresponding error code. 
%Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).
%\petr{Should also provide the reason why the bundle has failed?}
